Sofia Godovykh
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(psych)
library(magrittr)
library(tidyr)
library(memisc)
library(knitr)
library(RColorBrewer)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv('wineQualityWhites.csv', sep = ',')
wine <- na.omit(wine)
wine <- subset(wine, select = -c(X))
wine$quality_factor <-cut(wine$quality, c(0,5,6,9), 
                          labels = c("Poor", "Average", "Good"))
```

The white wine dataset contains information of acidity, sugar, pH level, and 
other chemical parameters. Every wine is graded by critics according to its 
quality. I want to find factors which contribute to wine quality and, 
specifically, find a difference in parameters between good and poor wines.

# Univariate Plots Section

There is a general information about the dataset

```{r echo=FALSE, Univariate_Plots}
summary(wine)
```

There are histograms of all parameters in a dateset.

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_Plots1}
w <- reshape2::melt(wine)
ggplot(w, aes(value)) + facet_wrap(~variable, scales = 'free_x', nrow = 4) +
  geom_histogram()
```

I will explore every feature in detail.

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_Plots_alcohol}
ggplot(aes(x=alcohol), data = wine) +
  geom_histogram(binwidth = 0.4) +
  coord_cartesian(xlim = c(8, 14))
```

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_lcohol_summary}
summary(wine$alcohol)
```

The distribution of alcohol shows that there are two peaks, a major one around 
9% and smaller one around 12%.

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_Plots_chlorides}
ggplot(aes(x=chlorides), data = wine) +
  geom_histogram(binwidth = 0.005) +
  coord_cartesian(xlim = c(0, 0.1))
```

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_chlorides_sum}
summary(wine$chlorides)
```

Chlorides form a bell-curved distribution with long right tail. It would be
interesting to know if different wine qualities distributed normally and which
wines form the tail. It would be interesting to have a look at the outliers,
because maximum value is way higher than median value and 3rd quantile.

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_Plots_tsd}
ggplot(aes(x=total.sulfur.dioxide), data = wine) +
  geom_histogram() +
  coord_cartesian(xlim = c(0, 300))
```

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_tsd_sum}
summary(wine$total.sulfur.dioxide)
```

Total sulfur dioxide histogram also resembles normal distribution.

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_Plots_va}
ggplot(aes(x=volatile.acidity), data = wine) +
  geom_histogram(binwidth = 0.02) +
  coord_cartesian(xlim = c(0.05, 0.5))
```

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_va_sum}
summary(wine$volatile.acidity)
```

This plot resembles normal distribution, but it has a long tail. Logarithmic 
scale will help us to get rid of it.

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_Plots_va_log}
ggplot(aes(x=volatile.acidity), data = wine) +
  geom_histogram(binwidth = 0.04) +
  coord_cartesian(xlim = c(0.1, 0.5)) + 
  scale_x_log10(breaks = seq(0.2, 0.5, 0.05))
```

This plot with log scale looks better. We can observe a peak around 0.25 g/L.

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_Plots_qf}
ggplot(aes(x=quality_factor), data = wine) +
  geom_histogram(binwidth = 0.04, stat = "count") +
  coord_cartesian()
```

There is a plot of number of wines in each quality category. I am interested 
how histogram of alcohol grouped by quality would look like. I explore it later,
in bivariate analysis section.

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_Plots_d}
ggplot(aes(x = density), data = wine) +
  geom_histogram() +
  coord_cartesian(xlim = c(0.985, 1.005)) 
```

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_d_sum}
summary(wine$density)
```

Density forms a normal distribution.

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_Plots_rs}
ggplot(aes(x = residual.sugar), data = wine) +
  geom_histogram(binwidth = 0.06) +
  coord_cartesian(xlim = c(0.75, 21)) + 
  scale_x_log10(breaks = seq(0, 25, 2))
```

It is distribution with two peaks: at 1.5 and 9.

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_Plots_ca}
ggplot(aes(x = citric.acid), data = wine) +
  geom_histogram(binwidth = 0.03) +
  coord_cartesian(xlim = c(0, 0.75))

```

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_ca_sum}
summary(wine$citric.acid)
```

There is a distribution with one major, 0.3; and minor peak, 0.5. Maximum value
is way higher median: 1.66, so I need to pay attention to outliers. 

# Univariate Analysis

### What is the structure of your dataset?

The dataset consists of numerical variables: fixed.acidity, volatile.acidity, 
citric.acid, residual.acid, chlorides, free.sulfur.dioxide, 
total.sulfur.dioxide, density, pH, sulphates, alcohol and quality.
Quality is an integer parameter, but it would be reasonable to treat it 
as a categorical one.


### What is/are the main feature(s) of interest in your dataset?

The main feature I want to explore is quality. Here I have found correlations 
between features to find out which ones would be useful.

```{r echo=FALSE, Univariate_Plots2, message = FALSE, warning = FALSE}
round(cor(wine[sapply(wine, function(x) !is.factor(x))]),2)
```

### What other features in the dataset do you think will help support your 
###investigation into your feature(s) of interest?\

There are negatively correlated features I want to pay attention to: 
volatile.acidity, chlorides, density, total.sulfur.dioxide and positevely 
correlated: alcohol. However, it is possible that I will find out something
interesting about citric acid and residual sugar.

### Did you create any new variables from existing variables in the dataset?

I created a categorical variable for wine quality: "Poor" for wines graded 
lower than 6, "Average" for 6 and "Good" for 7 and higher.

### Of the features you investigated, were there any unusual distributions? \

Most of the distributions are usual stadard distributions apart from
residual.sugar, and chlorides and volatile acidity with a long right tail.

# Bivariate Plots Section

I have already found which features are highly correlated with the target 
feature, quality. For more insights I built a matrix of boxplots

```{r echo=FALSE, Bivariate_Plots4}
wine %>%
  gather(-quality_factor, key = "var", value = "value") %>%
  ggplot(aes(y = value, x = quality_factor)) +
    geom_boxplot(alpha = 0.1) +
    facet_wrap(~ var, scales = "free")
```

There are some interesting insights: the range of alcohol in good wines is as 
wide as the range of all wines, while low-rated wines almost never have
alcohol higher than 12. The lowest graded wines have less residual sugar.
For more detailed analysis, I will build larger, separate plots for every 
interesting variable vs quality.

Scatterplot and boxplot of volatile.acidity and quality

```{r echo=FALSE, Bivariate_Plots5}
ggplot(aes(x = quality_factor, y = volatile.acidity), data = wine) + 
  geom_boxplot(alpha = 0.5, color = "blue") +
  geom_jitter(alpha = 0.08)
```

There were two peaks found in a previous section: at 0.15 and 0.25. This plot
explains these peaks: there is a lower boundary around 0.15 and median 
around 0.25.
Poor-quality wines in general have more volatile acidity. Good ones have less, 
and it has higher standard deviation. There is a main statistical information 
above.

```{r echo=FALSE, quality_factor_summary}
summary(wine$volatile.acidity)
summary(wine$volatile.acidity[wine$quality_factor == "Poor"])
summary(wine$volatile.acidity[wine$quality_factor == "Average"])
summary(wine$volatile.acidity[wine$quality_factor == "Good"])
```


Scatterplot and boxplot of chlorides and quality

```{r echo=FALSE, Bivariate_Plots6}
ggplot(aes(x = quality_factor, y = chlorides), data = wine) + 
  geom_boxplot(alpha = 0.5, color = "blue") +
  geom_jitter(alpha = 0.08)
```

Now it is clear that poor and average wines with extremely high chlorides level contribute to the long tail found in a previous section.
There is a tendency of lowering the chlorides level with raising of quality.

```{r echo=FALSE, chlorides_summary}
summary(wine$chlorides[wine$quality_factor == "Poor"])
summary(wine$chlorides[wine$quality_factor == "Average"])
summary(wine$chlorides[wine$quality_factor == "Good"])
```

Scatterplot and boxplot of density and quality

```{r echo=FALSE, Bivariate_Plots7}
ggplot(aes(x = quality_factor, y = density), data = wine) + 
  geom_boxplot(alpha = 0.5, color = "blue") +
  geom_jitter(alpha = 0.08)
```
There are a few outliers among average wines. There is a tendency of lowering density level with raising of quality.

```{r echo=FALSE, density_summary}
summary(wine$density[wine$quality_factor == "Poor"])
summary(wine$density[wine$quality_factor == "Average"])
summary(wine$density[wine$quality_factor == "Good"])
```

Scatterplot and boxplot of total.sulfur.dioxide and quality

```{r echo=FALSE, Bivariate_Plots8}
ggplot(aes(x = quality_factor, y = total.sulfur.dioxide), data = wine) + 
  geom_boxplot(alpha = 0.5, color = "blue") +
  geom_jitter(alpha = 0.08)
```

There is a tendency of lowering sulfur dioxide with raising of quality. 
Standard deviation is also becoming lower.

```{r echo=FALSE, total.sulfur.dioxide_summary}
summary(wine$total.sulfur.dioxide[wine$quality_factor == "Poor"])
summary(wine$total.sulfur.dioxide[wine$quality_factor == "Average"])
summary(wine$total.sulfur.dioxide[wine$quality_factor == "Good"])
```

Scatterplot and boxplot of citric.acid and quality

```{r echo=FALSE, Bivariate_Plots9}
ggplot(aes(x = quality_factor, y = citric.acid), data = wine) + 
  geom_boxplot(alpha = 0.5, color = "blue") +
  geom_jitter(alpha = 0.08)
```

I noticed an unusual peak around 0.5 in citric acid histogram. This plots shows
that there are lots of wines that have exactly 0.5 g/L of citric acid.
Median of good wines is slightly lower compared to other categories. Standard
deviation is getting lower with raising of quality.

```{r echo=FALSE, citric.acid_summary}
summary(wine$citric.acid[wine$quality_factor == "Poor"])
summary(wine$citric.acid[wine$quality_factor == "Average"])
summary(wine$citric.acid[wine$quality_factor == "Good"])
```

Scatterplot and boxplot of alcohol and quality

```{r echo=FALSE, Bivariate_Plots10}
ggplot(aes(x = quality_factor, y = alcohol), data = wine) + 
  geom_boxplot(alpha = 0.5, color = "blue") +
  geom_jitter(alpha = 0.08)
```

Good wines tend to contain more alcohol. It would be interesting to have a look
to a histogram of alcohol level separated by quality.

```{r echo=FALSE, Bivariate_Plots_qf_3}
ggplot(aes(x=alcohol, fill=quality_factor), data = wine) +
  geom_histogram(alpha = 1, binwidth = 1, position = "dodge", col = "black") +
  scale_fill_brewer(palette="Greys")
```

It is interesting how wines are distributed. Good wines frequently have higher 
alcohol level. Low quality wines have the lowest alcohol level.

```{r echo=FALSE, alcohol_summary}
summary(wine$alcohol[wine$quality_factor == "Poor"])
summary(wine$alcohol[wine$quality_factor == "Average"])
summary(wine$alcohol[wine$quality_factor == "Good"])
```

The correlation between density and alcohol is very high, as well as for 
density and residual.sugar. I will draw plot these plots.

```{r echo=FALSE, Bivariate_Plots11}
ggplot(aes(x = density, y = alcohol), data = wine) +
    geom_point(alpha = 0.1, na.rm = TRUE) +
    xlim(quantile(wine$density, 0.01),quantile(wine$density, 0.99))
```

Alcohol vs density plot shows that there are almost no wines with both high 
alcohol and high density, and no wines with less than average alcohol combined
with less than average density.

```{r echo=FALSE, Bivariate_Plots12}
ggplot(aes(x = density, y = residual.sugar), data = wine) +
    geom_point(alpha = 0.1, na.rm = TRUE) +
    xlim(quantile(wine$density, 0.01),quantile(wine$density, 0.99)) +
  ylim(quantile(wine$residual.sugar, 0.01),quantile(wine$residual.sugar, 0.99))
```

Interesting things with density vs residual sugar: there are almost no wines 
with both high density and low residual sugar, as well as there are no
wines with low density and high sugar.

# Bivariate Analysis

The most interesting thing I discovered is a relationship of alcohol vs quality.
There is a tendency of increasing an alcohol level with increasing its quality.
Density is the highest in wines with the lowest grade.

### Did you observe any interesting relationships between the other features \

There are two clusters in a plot with residual.sugar and density.

### What was the strongest relationship you found?

There is a very strong relationship between alcohol and quality: the highest 
rated wines have higher alcohol level.

# Multivariate Plots Section

I will build scatterplots with quality, alcohol and other features which could
be interesting.

```{r echo=FALSE, Multivariate_Plots1}
wine %>%
  gather(-fixed.acidity, -free.sulfur.dioxide, -pH, -sulphates, -alcohol, 
         -quality, -quality_factor, key = "var", value = "value") %>%
  ggplot(aes(y = value, x = quality_factor, color = alcohol)) +
    geom_point(alpha = 0.05, position = position_jitter(0.3)) +
    facet_wrap(~ var, scales = "free") +
    scale_color_distiller(palette="Spectral")
```

Then I want to have a closer look to features' relations.

There is a scatterplot with alcohol, sugar and quality.
High quality wines tend to have from 10 ABV to 14 ABV and, in most cases, 
no more than 10 gram per litre of residual sugar. Most of low quality wines
have a range of alcohol from 9 ABV to 10.5 ABV and sugar from 1 g/L to 18 g/L.

```{r echo=FALSE, Multivariate_Plots2}
ggplot(aes(y = residual.sugar, x = alcohol, color = quality_factor), 
       data = wine) +
    geom_point(alpha = 0.4, na.rm = TRUE, position = position_jitter(0.1)) +
    ylim(0,quantile(wine$residual.sugar, 0.99)) + 
    scale_color_brewer(palette="Set1")
```


Then there is a plot with citric acid, density and quality. High quality wines 
tend to have less density. The interesting thing is that there is a visibile 
peak of citric acid at 0.49 g/Ml.

```{r echo=FALSE, Multivariate_Plots3}
ggplot(aes(y = citric.acid, x = density, color = quality_factor), 
       data = wine) +
    geom_point(alpha = 0.3, na.rm = TRUE, position = position_jitter(0.0001)) +
    ylim(0,quantile(wine$citric.acid, 0.99)) + 
    xlim(quantile(wine$density, 0.01),quantile(wine$density, 0.99)) + 
    scale_color_brewer(palette="Set1")
```

There is a scatterplot with alcohol, volatile.acidity and quality.
While a majority of the highest quality wines have at least 11.5 ABV, there are
some wines with less alcohol. These wines have quite a low volatile acidity.
The least graded wines have a high acidity.

```{r echo=FALSE, Multivariate_Plots4}
ggplot(aes(y = volatile.acidity, x = alcohol, color = quality_factor), 
       data = wine) +
    geom_point(alpha = 0.3, na.rm = TRUE, position = position_jitter(0.3)) +
    ylim(quantile(wine$total.volatile.acidity, 0.01), 
         quantile(wine$volatile.acidity, 0.99)) + 
    xlim(quantile(wine$alcohol, 0.01),quantile(wine$alcohol, 0.96)) + 
    scale_color_brewer(palette="Set1")
```

Two last features I have not analyzed in deatils yet: total.sulfur.dioxide and 
chlorides. High-graded wines have fewer sulfur dioxide and chlorides.

```{r echo=FALSE, Multivariate_Plots5}
ggplot(aes(y = total.sulfur.dioxide, x = chlorides, color = quality_factor), 
       data = wine) +
    geom_point(alpha = 0.5, na.rm = TRUE, position = position_jitter(0.003)) +
    ylim(quantile(wine$total.sulfur.dioxide, 0.01), 
         quantile(wine$total.sulfur.dioxide, 0.99)) + 
    xlim(quantile(wine$chlorides, 0.01),quantile(wine$chlorides, 0.96)) + 
    scale_color_brewer(palette="Set1")
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Goods wines have more alcohol and less than average residual sugar. Low-ranked
wines have less alcohol and all range of residual sugar. Amount of sugar 
decreases with growth of quality. Good wines have less than average density. 
High rated wines have high level of alcohol and all range of volatile acidity.
Howewer, good wines with less alcohol have low acidity. High-graded wines have 
fewer sulfur dioxide and chlorides, while low-graded wines are high with both.
Turns out that alcohol level has the highest influence on wine's grade. There
are some exceptions from this observation: wines with high residual sugar and
low alcohol could be really good, as well as wines with low alcohol and low
volatile acidity. 

### Were there any interesting or surprising interactions between features?

Cholrides and total sulfur dioxide both negatively affect wine 
grade. The interesting thing is that all the good wines have less than average
of sulfur dioxide and cholrides at the same time.
Good wines tend to have more alcohol with one exception: combination of low ABV
and high residual sugar.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(x=alcohol, fill=quality_factor), data = wine) +
  geom_histogram(alpha = 0.7, binwidth = 1, position = "dodge", col = "black") + 
  labs(title = "Distribution of alcohol by volume", 
       x = "alcohol by volume, %", y = "count", 
       fill = "Wine quality") +
  geom_density(alpha = 0, aes(y = ..count.. , color = wine$quality_factor), 
               show.legend = FALSE) +
    geom_vline(data = wine[which(wine$quality_factor == 'Good'),], mapping = aes(xintercept = mean(alcohol)),  
               linetype="dashed", color = brewer.pal(4, "Set1")[3]) +
  geom_vline(data = wine[which(wine$quality_factor == 'Average'),], mapping = aes(xintercept = mean(alcohol)),  
               linetype="dashed", color = brewer.pal(4, "Set1")[2]) +
  geom_vline(data = wine[which(wine$quality_factor == 'Poor'),], mapping = aes(xintercept = mean(alcohol)),  
               linetype="dashed", color = brewer.pal(4, "Set1")[1]) +
  geom_vline(mapping = aes(xintercept = mean(wine$alcohol)),  
               linetype="dashed", color = "Black") +
  scale_fill_brewer(palette="Set1") +
  scale_color_brewer(palette="Set1")
```

### Description One

There is a histogram of alcohol by volume, split by wine quality. I noticed a 
strong correlation between quality and ABV and decided to look at it closer. I 
made three histograms with corresponding density plots and marked three means 
along with the general mean.
Low quality wines have a peak at 9.5% and a mean around 9.75%.
Average wines have a peak at 10.5%, almost like all the wines combined.
Good wines distribution does not have an obvious peak. The mean is at 11.4%. It 
is interesting that there are just a few low-graded wines with AVB higher than 
this value.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(y = volatile.acidity, x = alcohol, color = quality_factor, 
           alpha = 0.8), data = wine) +
  geom_point(alpha = 0.25, na.rm = TRUE, position=position_jitter(0.05))+
  ylim(quantile(wine$volatile.acidity, 0.01), 
         quantile(wine$volatile.acidity, 0.99)) + 
  xlim(quantile(wine$alcohol, 0.01),quantile(wine$alcohol, 0.96)) + 
  labs(title = "Affect of volatile acidity and alcohol on wine quality", 
       x = "alcohol by volume, %", y = "volatile acidity, g/L", 
       color = "Wine quality") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  scale_color_brewer(palette="Set1") +
  stat_ellipse(aes(color = quality_factor), geom = "path", level=0.8, 
               alpha = 1, show.legend = FALSE, na.rm = TRUE)
```

### Description Two

This plot describes volatile acidity, alcohol and quality. Ellipses cover 80% 
of samples. According to this plot, poor wines form a cluster with alcohol from
8.5% to 11% and acidity from 0.15 g/L to 0.43 g/L.
Average wines ellipse is wider: it has alcohol range 8.5%-12% and acidity from
0.15 g/L to 0.35 g/L. 
Good wines stay apart from others: the center of the cluster is significantly on the right at the alcohol scale. Alcohol varies from 9.5% to 13%, acidity varies 
from 0.12 to 0.375.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(y = total.sulfur.dioxide, x = chlorides, color = quality_factor, 
           alpha = 1), data = wine) +
  geom_point(alpha = 0.2, na.rm = TRUE, position=position_jitter(0.001)) +
  #geom_smooth(aes(color = quality_factor, fill = quality_factor), 
              #show.legend = FALSE, na.rm = TRUE, method = "auto") +
  stat_ellipse(aes(color = quality_factor), geom = "path", level=0.9, 
               alpha = 1, show.legend = FALSE, na.rm = TRUE) +
  ylim(quantile(wine$total.sulfur.dioxide, 0.01), 
         quantile(wine$total.sulfur.dioxide, 0.99)) + 
  xlim(quantile(wine$chlorides, 0.01),quantile(wine$chlorides, 0.96)) + 
  labs(title = "Affect of total sulfur dioxide and chlorides on wine quality", 
       x = "chlorides, g/L", y = "total sulfur dioxide, mg/L", 
       color = "Wine quality") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  scale_color_brewer(palette="Set1") 
```

### Description Three

This plot describes affect of chlorides and sulfur dioxide on wine quality. 
Ellipses represent 90% of samples. Good wines have lower chlorides and sulfur 
dioxide, poor wines contains more chlorides and sulfur dioxide. Despite of this
trend, the ellipses share quite a wide common area.

------

# Reflection

I explored how different parameters are associated with wine quality. I split
wine quality for three categories: poor, average and good and perform an 
analysis. In general, good wines contain more alcohol and volatile acidity, 
less chlorides, sulfur dioxide, density and sugar.
I was challenged with building a linear model for quality classification. I used
alcohol, volatile acidity, chlorides, sulfur dioxide, density and sugar as 
features to predict a quality class. The accuracy of the model was only 60%. For
better results it would be useful to tune the model better or use more advanced 
method.

# Resources

https://stackoverflow.com  
https://www.statmethods.net  
http://www.sthda.com/  
http://ggplot2.tidyverse.org/index.html  

